---
title: "NBA Sports Betting and Predictions"
output: 
  flexdashboard::flex_dashboard:
    theme: 
      bootswatch: zephyr
    orientation: columns
    vertical_layout: fill
    source_code: embed
---

<style>
.chart-title {  /* chart_title  */
   font-size: 16px;
  }
body{ /* Normal  */
      font-size: 14px;
  }
</style>

```{r setup, include=FALSE}
library(flexdashboard)
library(pacman)

p_load(tidyverse, maps, viridis, ggplot2, plotly, htmltab, readr, leaflet, scales, plyr, reshape2, ggcorrplot, gridExtra)
```

Introduction
===

Column {data-width = 500} 
---

### A Brief History & Background Information

In 1931, Nevada becomes the first place in the United States where
gambling and betting on sports is legal and it remains the only place to do
so for decades. In 1975, the first true "sportsbook casino" is built in Las Vegas.
In 2018, the United States Supreme Court ruled that each state can freely establish their own sports gambling laws. Delaware stepped up first, joining Nevada, and by the end of the year, so did New Jersey, Rhode Island, Mississippi, West Virginia, New Mexico, Arkansas, and Pennsylvania. Also in 2018, popular online sportsbook DraftKings establishes itself in New Jersey. At the time, DraftKings was the first online sportsbook outside of Nevada. Currently, at time of writing, sports betting is legal with regulated industries in thirty-six states and Washington D.C. Two more states have anticipated launches in 2024 with five more reportedly close behind.
Popular online sportsbooks across the country include DraftKings, FanDuel, BetMGM, and Bet365.

### Why Analyze?

The idea behind this data set is to discover any trends between pre-game betting values, such as moneylines or spreads, and overall outcome of the game. Obviously, sportsbooks want to make the most money they can off of betters and betting is never a sure thing, but with a little more information, maybe those betters can place more informed bets with a lower risk.

Column {data-width = 500} 
---

### Goals

The goal of this project is to create prediction models that predict the outcome of the game based on pre-game statistics, such as moneyline and spread. If possible, then the goal becomes finding out how accurately this model can predict the final outcome. Does this vary by home team and away team? Finally, to create an interactive platform for users to input pre-game statistics and give them an output.


Data Introduction
===

Column {.tabset data-width=550}
---

### Player Data

```{r}
playerdata <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_all_players_2021-22.csv")
teamdata <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/teamdata.csv")

gamelogs2122 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_gamelogs_2021-22.csv")

##odds1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_fpts.csv")
##teamodds1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_fpts_team.csv")

ftts1212 <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/odds_ftts.csv")

outputfpts <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_fpts.csv")
outputftts <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_ftts.csv")

outputwintip <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/output_win_tip.csv")

samplemoneyline <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_20221018_moneyline.csv")

samplespread <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/nba_20221018_spread.csv")

moneylines <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/allmoneylines.csv")

spreads <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/allspreads.csv")

tipoffscorefirst <- read.csv("C:/Users/margo/OneDrive/Documents/MTH 369/nbabettinglines/tipoffscorefirst.csv")
```

```{r}
playerdata <- playerdata[ , names(playerdata) %in% c("PLAYER_NAME", "TEAM_ABBREVIATION", "AGE", "PLAYER_HEIGHT_INCHES", "PLAYER_WEIGHT", "GP", "PTS", "REB", "AST", "NET_RATING", "OREB_PCT", "DREB_PCT", "USG_PCT", "TS_PCT", "AST_PCT", "SEASON_YEAR")]

teamdata <- teamdata[ , names(teamdata) %in% c("SEASON_YEAR", "TEAM_CITY", "TEAM_NAME", "TEAM_ABBREVIATION", "TEAM_CONFERENCE", "TEAM_DIVISION", "MIN_YEAR", "MAX_YEAR", "TEAM_FULL_NAME")]

samplemoneyline <- samplemoneyline[ , names(samplemoneyline) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line", "away_outcome", "home_outcome")]

samplespread <- samplespread[ , names(samplespread) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line", "away_outcome", "home_outcome")]

moneylines <- moneylines[ , names(moneylines) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_line", "home_line", "away_outcome", "home_outcome")]

spreads <- spreads[ , names(spreads) %in% c("game_id", "Date", "bet_type", "away_Team", "home_Team", "away_spread", "home_spread", "away_outcome", "home_outcome")]
```

```{r}
DT::datatable(playerdata, rownames = FALSE, 
              colnames = c("Player Name", "Team Abbreviation", "Age", "Height (in.)", "Weight", "Games Played", "Points", "Rebounds", "Assists", "Net Rating", "Offensive Rebound %", "Defensive Rebound %", "Usage %", "Total Scoring %", "Assist %", "Season"))
```

### Team Data
```{r}
DT::datatable(teamdata, rownames = FALSE, 
              colnames = c("Season", "City", "Name", "Abbreviation", "Conference", "Division", "Min Year", "Max Year", "Team Full Name"))
```

### Moneyline Example
```{r}
DT::datatable(samplemoneyline, rownames = FALSE, 
              colnames = c("Teams", "Date", "Bet Type", "Away Team", "Home Team", "Away Line", "Home Line", "Away Team Outcome", "Home Team Outcome"))
```

### Spread Example
```{r}
DT::datatable(samplespread, rownames = FALSE, 
              colnames = c("Teams", "Date", "Bet Type", "Away Team", "Home Team", "Away Spread", "Home Spread", "Away Team Outcome", "Home Team Outcome"))
```

### Variable Descriptions

**Player Data**

Player Name: Name of NBA player

Team Abbreviation: A unique three letter code that defines the team name

Age: Age of player, in years

Height: Height of player, in inches

Weight: Weight of player, in pounds

Games Played: How many games this player has played in over the course of the entire NBA season

Points: How many total points this player scored across the entire season

Rebounds: How many total rebounds this player accumulated over the season

Assists: How many total assists this player had over the entire season

Net Rating: Point differential 100 possessions

Offensive Rebound %: Percentage of available offensive rebounds a player obtains while on the floor

Defensive Rebound %: Percentage of available defensive rebounds a player obtains while on the floor

Usage %: Percentage of team plays used by a player while they are on the floor

True Shooting %: Shooting percentage that factors in the value of three-point field goals and free throws, in addition to two-point field goals

Assist %: Percentage of a team's assists a player has while on the court

Season: Years that the basketball games were played


**Team Data**

Season: Years that the basketball games were played

City: The city in which the team plays in or near

Name: The mascot or team descriptor

Abbreviation: A unique three letter code that defines the team name

Conference: East or West, a location based descriptor that separates the teams into two groups

Division: blank, a more specific location based descriptor that divides each conference into

Min Year: The first year that the team participated in the NBA

Max Year: The most recent year that the team participated in the NBA


**Moneyline Example**

Teams: In a x_y format, each variable is a team abbreviation, describing who played the game

Date: In a YYYYMMDD format, the date in which the game took place

Bet Type: The type of bet being tracked (in this case, moneyline)

Away Team: The team in which is the visitor

Home Team: The team that is hosting the game in their city and arena

Away Line: Average moneyline between sportsbooks for the visiting team. A positive value denotes the "underdog" and a negative value indicates the favorite to win

Home Line: Average moneyline between sportsbooks for the hosting team. A positive value denotes the "underdog" and a negative value indicates the favorite to win


**Spread Example**

Teams: In a x_y format, each variable is a team abbreviation, describing who played the game

Date: In a YYYYMMDD format, the date in which the game took place

Bet Type: The type of bet being tracked (in this case, spread)

Away Team: The team in which is the visitor

Home Team: The team that is hosting the game in their city and arena

Away Spread: If negative, the value the visiting team must win by. If positive, if the team loses the game by less than the number or if the team wins overall the bet will win

Home Spread: If negative, the value the hosting team must win by. If positive, if the team loses the game by less than the number or if the team wins overall the bet will win



Sports Betting Basics
===

Column {.tabset data-width=550}
---

### Moneyline & Spread

In basketball, there are many types of bets a potential better could place. Bets are available for player stats, such as points scored, rebounds, or assists and team stats such as divisional winners or future award winners. Ultimately, the most popular and simpler sports bets utilized for basketball are the moneyline and the spread.

The **moneyline** is a bet placed on either team to win the match and thus, has two outcomes. The team favored to win will have a negative moneyline value and the team who is expected to lose will have a positive moneyline value. Suppose the Cleveland Cavaliers have a +150 moneyline and the Indiana Pacers have a -150 moneyline. This means that if you bet $100 on the Cavaliers, you will earn $150. However, a $150 bet on the Pacers will only earn $100.

The **spread** is common in the NBA because when you bet against the spread there are more than the two outcomes found in the moneyline. The spread is a smaller value that depicts point favorites. Let's take the example from above, except now the Cavaliers are -7 against the spread. This means that they must win by 8 or more points. If the Pacers lose by 6 points or fewer or win the game, they cover the spread.

### Tip-Off & Score First

In basketball, bets can be placed on which player will score the first basket, but that's not useful for this analysis. In this project, the tip-off true/false shows if the home team won the tip-off (has first possession of the ball) and the score first true/false shows if the home team scored the first basket.

Once again, this is being used to find a correlation (if any) between winning the tip-off or scoring the first basket, with moneylines, spreads, or winning the game.


Graphical Displays
===

Column {.tabset data-width=550}
---

### Moneylines

```{r, figures-side, fig.show="hold", out.width="50%"}
par(mfrow = c(2,2))

ggplot(moneylines, aes(x = away_line)) + geom_histogram(bins = 20, fill = "cornflowerblue") + labs(title = "Distribution of Away Moneylines", x = "Away Line", y = "Count")

ggplot(moneylines, aes(x = home_line)) + geom_histogram(bins = 25, fill = "cornflowerblue") + labs(title = "Distribution of Home Moneylines", x = "Home Line", y = "Count")

ggplot(moneylines, aes(x = as.factor(away_outcome), y = away_line)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Away Moneyline vs Outcome", x = "Win or Loss", y = "Moneyline")

ggplot(moneylines, aes(x = as.factor(home_outcome), y = home_line)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Home Moneyline vs Outcome", x = "Win or Loss", y = "Moneyline")
```

### Spreads

```{r, figures-side2, fig.show="hold", out.width="50%"}
par(mfrow = c(2,2))

ggplot(spreads, aes(x = away_spread)) + geom_histogram(bins = 20, fill = "cornflowerblue") + labs(title = "Distribution of Away Spreads", x = "Away Spread", y = "Count")

ggplot(spreads, aes(x = home_spread)) + geom_histogram(bins = 25, fill = "cornflowerblue") + labs(title = "Distribution of Home Spreads", x = "Home Spread", y = "Count")

ggplot(spreads, aes(x = as.factor(away_outcome), y = away_spread)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Away Spreads vs Outcome", x = "Win or Loss", y = "Away Spread")

ggplot(spreads, aes(x = as.factor(home_outcome), y = home_spread)) + 
    geom_boxplot(fill = "royalblue1", alpha = 0.5) + labs(title = "Home Spreads vs Outcome", x = "Win or Loss", y = "Home Spread")
```


### Moneyline vs Spread

```{r, figures-side3, fig.show="hold", out.width="50%"}
moneyspread <- moneylines %>% left_join(spreads, by = "game_id", relationship = "many-to-many")

moneyspread <- moneyspread[ , names(moneyspread) %in% c("game_id", "Date.x", "away_Team.x", "home_Team.x", "away_line", "home_line", "away_spread", "home_spread", "away_outcome.y", "home_outcome.y")]

names(moneyspread)[2] <- 'Date'
names(moneyspread)[3] <- 'away_Team'
names(moneyspread)[4] <- 'home_Team'
names(moneyspread)[9] <- 'away_outcome'
names(moneyspread)[10] <- 'home_outcome'


ggplot(moneyspread, aes(x = away_line, y = away_spread)) +
  geom_point(color = "cornflowerblue") + labs(title = "Away Line vs Spread", x = "Moneyline", y = "Spread")

ggplot(moneyspread, aes(x = away_line, y = away_spread, color = away_outcome)) +
  geom_point() + theme(text = element_text(size = 20)) + labs(title = "With Outcome", x = "Moneyline", y = "Spread")

ggplot(moneyspread, aes(x = home_line, y = home_spread)) +
  geom_point(color = "cornflowerblue") + labs(title = "Home Line vs Spread", x = "Moneyline", y = "Spread")

ggplot(moneyspread, aes(x = home_line, y = home_spread, color = home_outcome)) +
  geom_point() + theme(text = element_text(size = 20)) + labs(title = "With Outcome", x = "Moneyline", y = "Spread")
```

### Win Tip

```{r}
ggplot(tipoffscorefirst, aes(x = home_outcome, fill = home_won_tip)) + 
  geom_bar() + labs(title = "Home Outcome vs Tip-Off", x = "Outcome", y = "Count")
```

### Score First

```{r}
ggplot(tipoffscorefirst, aes(x = home_outcome, fill = home_score_first)) + 
  geom_bar()+ labs(title = "Home Outcome vs First Score", x = "Outcome", y = "Count")
```

Basic Prediction
===

### Home Predictions

```{r}
attach(moneyspread)
home.ms <- moneyspread[ , names(moneyspread) %in% c("home_line", "home_spread", "home_outcome")]
home.ms$home_outcome <- as.factor(home.ms$home_outcome)

train <- sample(c(1:497), 397)
train <- home.ms[train,]

test <- sample(c(1:497), 100)
test <- home.ms[test,]
```

```{r}
library(naivebayes)
fit <- naive_bayes(home_outcome ~ ., data = home.ms[1:397,])
summary(fit)

preds <- predict(fit, home.ms[1:100,], interval = "confidence")
preds
```

```{r}
library(randomForest)
rf.hms <- randomForest(home_outcome ~ ., data = home.ms[1:397,], mtry = 4, importance = TRUE)
rf.hms

importance(rf.hms)
varImpPlot(rf.hms)

pred.rfhms <- predict(rf.hms, newdata = home.ms[1:100,])
pred.rfhms
```

Interactive Prediction
===

### Home Outcome Prediction

```{r}
home.ms$home_outcome <- as.factor(ifelse(home.ms$home_outcome == 'W', '1', '0'))

library(shiny)

model <- lm(home_outcome ~ home_line + home_spread, data = home.ms)

ui <- fluidPage(
  titlePanel("Predicting Home Win/Loss Outcome"),
  sidebarLayout(
    sidebarPanel(
      numericInput("home line input", "Enter home line:", value = 100),
      numericInput("home spread input", "Enter home spread:", value = 8),
      actionButton("predict_button", "Predict"),
      hr(),
      h4("Model Coefficients:"),
      verbatimTextOutput("coefficients_text")
    ),
    mainPanel(
      style = "background-color: #C0B8B8;",
      plotOutput("regression_plot"),
      h4("Predicted Outcome:"),
      verbatimTextOutput("predicted_outcome")
    )
  )
)

server <- function(input, output) {
  predict_outcome <- function(home_line, home_spread) {
    new_data <- data.frame(home_line = home_line, home_spread = home_spread)
    predicted_outcome <- predict(model, newdata = new_data)
    return(predicted_outcome)
  }
  
output$regression_plot <- renderPlot({
    ggplot(data, aes(x = home_line, y = home_outcome)) +
      geom_point(aes(color = "home_line"), size = 3) +
      geom_smooth(method = "lm", formula = home_outcome ~ home_line + home_spread, se = FALSE) +
      geom_point(aes(x = home_spread, color = "home_spread"), size = 3) +
      labs(title = "Multiple Linear Regression",
           x = "home line and home spread",
           y = "home outcome",
           color = "Variable") +
      theme_minimal()
  })

output$coefficients_text <- renderText({
  paste("Intercept:", round(coef(model)[1], 4),
        "\nx2 Coefficient:", round(coef(model)[2], 4),
        "\nx5 Coefficient:", round(coef(model)[3], 4))
})

observeEvent(input$predict_button, {
  homeline_input <- input$homeline_input
  homespread_input <- input$homespread_input
  predicted_outcome <- predict_outcome(homeline_input, homespread_input)
  output$predicted_outcome <- renderText({
    paste("Home Outcome:", round(predicted_outcome, 2))
  })
})
}

shinyApp(ui, server)
```


Analysis / Conclusion
===

In conclusion, I was able to create several prediction models and even an interactive platform where the user can input their own values. However, like sports betting, nothing is guaranteed and the models do make mistakes. More often than not, though, the predictions are correct.